# Introduction {#introduction}
\chaptermark{Introduction}
\minitoc

<!--   
Structure of this chapter:
  - General introduction to MV extremes
  - Data write up, France rainfall dataset
  - Outline of this report
-->


\noindent \todo{General introduction to this report.}

## French precipitation maxima data

The French precipitation maxima dataset consists of the weekly maxima of hourly precipitation at 92 weather stations in France. The measurements are taken during the autumn season (September-November) between 1993 and 2011. The data was collected by Météo-France, the French meteorological service, and is available from the homepage of the second author of @bernardClusteringMaximaSpatial2013 at [https://www.lsce.ipsl.fr/Phocea/Pisp/visu.php?id=109&uid=naveau](https://www.lsce.ipsl.fr/Phocea/Pisp/visu.php?id=109&uid=naveau).

```{r load-france-data, include=FALSE}
# load the France rainfall dataset
load(file = "Code/BernardEtAl2013/MaxPrecipFallFrance.RData")

frx <- MaxPrecipFallFrance$precip
frLong <- MaxPrecipFallFrance$longitudes
frLat <- MaxPrecipFallFrance$latitudes
```

The weather stations provide a fairly complete and homogeneous coverage of France, as illustrated in Figure \@ref(fig:fr-data) (left). At each station, we have a time series of length $T=228$. Figure \@ref(fig:fr-data) (right) shows one such the time series, corresponding to the weather station at Château-Gaillard, which is approximately 50km north-east of Lyon and marked in red in Figure \@ref(fig:fr-data) (left). 

```{r fr-data, echo=FALSE, fig.align='center', fig.cap='The locations of the 92 French weather stations, with Château-Gaillard marked in red (left) and the time series of weekly maxima of hourly precipitation recorded at Château-Gaillard (right).', fig.scap='France precipitation maxima data.', out.width="100%"}

par(mfrow=c(1,2), mar=c(5, 4, 4, 2))

# plot a map of all the stations
neighbors <-  c("France","Spain","Germany","Belgium","Italy","Switzerland","Luxembourg","Netherland","Monaco","Andorra")
bgOcean <- "white"
bgFrance <- "grey85"
map("worldHires", neighbors, col = "grey95", fill = T, bg = bgOcean, xlim = c(-5.0,9.5), ylim = c(41.5,51), mar=c(0, 0, 0, 5))
map("worldHires", "France", col = bgFrance, fill = T, add = T)
map.axes(xaxt='n', yaxt='n', ann=FALSE)
grid()
points(frLong[-1], frLat[-1], col = "black", pch = 20, cex = 0.8) # plot all stations except Château-Gaillard in black
points(frLong[1], frLat[1], col = "red", pch = 20, cex = 0.8) # plot Château-Gaillard in red

# plot time series of Château-Gaillard (the first station in the dataset)
y <- as.numeric(frx[,1])
tmax <- length(y)
plot(1:tmax, y, pch = 20, cex = 0.8, xlab = "Index", ylab = "Maximum hourly precipitation (mm)", cex.lab = 0.7, cex.axis = 0.7, mgp=c(1.5,0.5,0))
```

## Report outline

\todo{Give an outline of how the report is organised}
\begin{itemize}
\item Chapter 2 - mathematical background for EV theory
\item Chapter 3 - clustering
\item Chapter 4 - Causal inference
\item Chapter 5 - PCA
\item Chapter 6 - General conclusions/discussion
\end{itemize}


